### 2022-10-24 dissertation/chapters/03-heuristics.tex
🚧 (chapter 3): Just before DE

      \item \textbf{Section~\ref{sec:heuristics:what_is_a_heuristic}} provides background information on the origins and definition the term \index{heuristic}\textit{heuristic}. It is shown that \index{heuristic}heuristics are a class of algorithms that are used to solve optimisation problems.

@@

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''.~\citeauthor{ref:romanycia:1985}~\cite{ref:romanycia:1985} provides a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \acf{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \index{hmeta-euristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \index{meta-heuristic}meta-heuristics make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. Each of these are presented in detail in the following sections.

@@

Although \acs{GD} \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \acp{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most whidely used. \Acs{GD} heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics and other second-order derivative methods such as Newtons' method. There are many variants of \acs{GD} \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acs{GD} framework called \acl{BP}.

@@

\noindent
The update step for \acs{GD} can be formulated as is shown in Equation~\eqref{eq:heuristics:gd:update_step_part_1} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_1} above, the delta weight $\Delta w$, at index $i$ and timestep $t$ is given by Equation~\eqref{eq:heuristics:gd:update_step_part_2} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_2} above, $\eta$ is the learning rate. The learning rate is an important hyper-parameter in gradient-based \index{heuristic}heuristic as it controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given by Equation~\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
In the context of training \textit{shallow} \acp{FFNN}, using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer. The algorithm that propagates the error signal backwardds is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\epsilon$ with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007}describes the \acs{BP} process in two steps:

      \item \textbf{Backward Propagation}: During this phase the error signal that is calculated as is shown above in Equations~\eqref{eq:heuristics:gd:update_step_part_1}, ~\eqref{eq:heuristics:gd:update_step_part_2} and~\eqref{eq:heuristics:gd:update_step_part_3} is propagated backwards from the output layer, through the hidden layer, to the input layer of the \acs{FFNN}. Weights are then adjusted as functions of the backpropagated error signal.
\end{itemize}

\noindent
The pseudo-code implementation for the generic \acs{GD} algorithm is taken from~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:gd} below.

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the generic \acf{GD} heuristic.}
      \label{algo:heuristics:gd:gd}
      \begin{algorithmic}
            \State Initialise the \acs{FFNN} weights and biases, the learning rate $\eta$ and the timestep/epoch $T=0$;
            \While{stopping conditions are not met}
            \State Let $\epsilon_{T} = 0$;
            \For{each training pattern p}
            \State Do the feedforward pass phase to calculate the output of the hidden layer $y_{j,p}$ ($\forall j = 1, \dots, J$), and the output of the output layer $o_{k,p}$ ($\forall k = 1, \dots, K$);
            \State Compute the output error signals $\delta_{o_{k,p}}$ and the hidden layer error signals $\delta_{y_{j,p}}$;
            \State Adjust the weights of the output layer $w_{kj}$, and the weights of the hidden layer $v_{ji}$ (backpropagation of errors);
            \State $\epsilon_{T}$ += $[\epsilon_{T_{p}} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$; (assuming the use of \acs{SSE} as the loss function)
            \EndFor
            \State $T += 1$;
            \EndWhile
      \end{algorithmic}
\end{algorithm}

\noindent
In order to provide a simplified frame of reference from which the gradient-based heuristics are compared, consider a simplification of the update-step for \acs{SGD} as presented in Equation~\eqref{eq:heuristics:gd:update_step_part_1}. This simplification is depicted in Equation~\eqref{eq:heuristics:gd:sgd} below.

\begin{equation}
      \label{eq:heuristics:gd:sgd}
      \begin{split}
            w = w - \eta \nabla_{w}\mathcal{E}(w)
      \end{split}
\end{equation}

\noindent
In Equation~\eqref{eq:heuristics:gd:sgd} above, $w$ refers to the weights of the \acs{FFNN}, $\eta$ refers to the learning rate as before and $\nabla_{w}\mathcal{E}(w)$ refers to the gradient of the loss function $\mathcal{E}$, w.r.t the output of the \acs{FFNN} as a result of its weights. Note that all subscripts w.r.t. layers and training patterns are omitted for convenience.

@@

This section shines light on the algorithmic implementation of \ac{BP} with specific context to stochastic and batch training. The implementation of \acs{GD} using stochastic training is referred to as \acf{SGD}.

\acs{SGD} was one of the first widely used heuristics to train \ac{FFNN}, however, it is not without flaws. With stochastic training, only one training pattern is presented at each iteration/epoch. As such, weight updates are done with high variance and noise~\cite{ref:ruder:2016}. An illustration of the the fluctuations caused by \ac{SGD} during training is given in Figure~\ref{fig:heuristics:gd:sgd}.

@@

A compromise to overshooting minima that \acs{SGD} suffers from, is to include mini-batches of training patterns where a small number of training patterns are presented to the \acs{FFNN} at once. This is referred to as \textit{mini-batch} training. Input patterns from mini-batches are approximations of the total population of training patterns. According to~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} this has two main advantages:

@@

      \item The learning rate that has been introduced so far is applied to all elements of the weight vector updates. If the training data is sparse and the features have very different frequencies, an equal update to all weights elements is inefficient. Larger weight updates are required for less frequently occurring features. Inefficient updates of all elements in the weight vectors contribute to the credit assignment problem~\cite{ref:rumelhart:1986}, common to \acs{GD} variants.

@@

Research shows that \ac{SGD} has difficulty navigating ravines~\cite{ref:sutton:1986}. Ravines are areas where the surface curves much more steeply in one dimension than in another. These ravines are common around local minima. As such, \acs{SGD} is shown to oscillate across the slopes of the ravine while only making minor progress towards the local minima.

\Acs{Momentum} ~\cite{ref:qian:1999} is a variant of the \acs{SGD} that helps accelerate \ac{SGD} in the relevant direction, dampening oscillations.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$, of the weight update vector of the previous timestep to the current weight update vector. An illustration of \acs{SGD} with and without momentum is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

@@

\noindent
The accumulation of momentum is presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1} below, while the update step is then amended in Equation~\ref{eq:heuristics:gd:momentum_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_1}
      \begin{split}
            v_{t} = \alpha v_{t-1} + \eta \nabla_{w}\mathcal{E}(w)
      \end{split}
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_2}
      \begin{split}
            w = w - v_{t}
      \end{split}
\end{equation}

@@

\Acf{NAG} is a variant of the \acs{Momentum} heuristic developed by~\citeauthor{ref:sutskever:2013-2}~\cite{ref:sutskever:2013-2} and is inspired from~\citeauthor{ref:nesterov:1983}~\cite{ref:nesterov:1983} work on optimising convex functions. \ac{NAG} provides an improvement to the momentum accumulation term by providing a look-ahead term that better refines the weight update step.

@@

\noindent
The difference in the update step for \acs{Momentum} and the update step for \ac{NAG} as developed by~\citeauthor{ref:sutskever:2013-2}~\cite{ref:sutskever:2013-2} is described by~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} as follows. \ac{Momentum} first calculates the current gradient, represented by the small blue vector in Figure~\ref{fig:heuristics:gd:nag} and then takes a large step in the direction of the updated accumulated gradient, presented by the big blue vector. In contrast, \acs{NAG} first takes a big step in the direction of the previous accumulated gradient presented by the brown vector. At this point the gradient is measured and \acs{NAG} makes a correction represented by the red vector. The complete update step is represented by the green vector. By anticipating approximate future positions of the weights, the weight update step as defined by \ac{NAG} controls the optimisation process from going too fast and results in increased responsiveness~\cite{ref:bengio:2013}. The \acs{NAG} update steps are given in Equations~\eqref{eq:heuristics:gd:nag_part_1} and~\eqref{eq:heuristics:gd:nag_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:nag_part_1}
      \begin{split}
            v_{t} = \alpha v_{t-1} + \eta \nabla_{w}\mathcal{E}(w - \alpha v_{t-1})
      \end{split}
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:nag_part_2}
      \begin{split}
            w = w - v_{t}
      \end{split}
\end{equation}

@@

\Acf{Adagrad} is an adaptation of \ac{SGD} which implements a learning rate for every element in the weight vector and is developed by Duchi et al.~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \acs{Adagrad} adapts the learning rate to the elements of the weight vector, by  performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason, \acs{Adagrad} is well suited for dealing with situations where training data is sparse.

In the \acs{GD} variants presented thus far, the weight updates have been applied by making use of the same learning rate $\eta$ for all elements of the weight update vector. \Acs{Adagrad} uses a different learning rate for every weight element $w_{i}$, at every timestep $t$. The weight update step is presented in Equation~\eqref{eq:heuristics:gd:adagrad_part_1} below.

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_1}
      \begin{split}
            w_{t+1,i} = w_{t,i} - \frac{\eta}{\sqrt{G_{t,ii} + \epsilon}}.\nabla_{w_{i}}\mathcal{E}(w_{t,i})
      \end{split}
\end{equation}

\noindent
In Equation~\eqref{eq:heuristics:gd:adagrad_part_1} above, $w_{i}$ is the $i$-th element of the weight vector, $G_{t,ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$ is the sum of the squared gradients w.r.t. $w_{i}$, up to timestep $t$ and $\epsilon$ is a smoothing term that avoids division by zero and is set to an insignificant value such as $1 \times 10^{-8}$. Since $G_{t}$ contains the sum of the squares of the gradients along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product, represented by $\odot$, between $G_{t}$ and $\nabla_{w_{i}}\mathcal{E}(w_{i})$. The simplified update step as implemented by \acs{Adagrad} is given in Equation~\eqref{eq:heuristics:gd:adagrad_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_2}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{\sqrt{G_{t} + \epsilon}} \odot \nabla_{w_{i}}\mathcal{E}(w_{t})
      \end{split}
\end{equation}

\noindent
\Ac{Adagrad}'s main benefit is that it does not require manual tuning of the learning rate $\eta$. However, a problem with \ac{Adagrad} is in the accumulation of squared gradients in the denominator, represented by $G_{t}$. Since every term that is added is positive, $G_{t}$ keeps growing, leading to a situation where the learning rate shrinks to the point that the heuristic is no longer able to learn.

@@

Zeiler~\cite{ref:zeiler:2012} presents \ac{Adadelta}, an improvement of \acs{Adagrad} that eliminates the accumulation of squared gradients in the denominator.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \ac{Adagrad}, \ac{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. In order to simplify the notation of update steps, let $g_{t} = \nabla_{w_{i}}\mathcal{E}(w_{t})$. The moving average of squared gradients at timestep $t$, denoted by $E[g^{2}]_{t}$  depends on a fraction $\alpha$ of the previous average of squared gradients and the current gradient, similar to the update step for \acs{Momentum}. Similar to \acs{Momentum}, $\alpha$ is set to 0.9. The update step for $E[g^{2}]_{t}$ is presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_1} below.

@@

\noindent
To illustrate the difference between \acs{Adagrad} and \acs{Adadelta}, the \ac{SGD} update step as presented in~\eqref{eq:heuristics:gd:sgd} is rewritten in terms of the weight update vector $\Delta w_{t}$ and gradient $g_{t}$ and is presented in Equations~\ref{eq:heuristics:gd:adadelta_part_2} and~\eqref{eq:heuristics:gd:adadelta_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_2}
      \begin{split}
            \Delta w_{t} = -\eta g_{t,i}
      \end{split}
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_3}
      \begin{split}
            w_{t+1} = w_{t} + \Delta w_{t}
      \end{split}
\end{equation}

\noindent
The weight update vector for \ac{Adagrad} is then redefined in Equation~\eqref{eq:heuristics:gd:adagrad_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_2_new}
      \begin{split}
            \Delta w_{t} = - \frac{\eta}{\sqrt{G_{t} + \epsilon}} \odot g_{t}
      \end{split}
\end{equation}

\noindent
The diagonal matrix $G_{t}$, implemented by \ac{Adagrad}, is then replaced with the moving average of squared gradients as presented in Equation~\ref{eq:heuristics:gd:adadelta_part_4} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_4}
      \begin{split}
            \Delta w_{t} = - \frac{\eta}{\sqrt{E[g^{2}]_{t} + \epsilon}} g_{t}
      \end{split}
\end{equation}

\noindent
Since the denominator is the \acf{RMS} error criterion of the gradient, the criteria short-hand notation is used as is depicted in Equation~\eqref{eq:heuristics:gd:adadelta_part_5} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_5}
      \begin{split}
            \Delta w_{t} = - \frac{\eta}{RMS[g]_{t}} g_{t}
      \end{split}
\end{equation}

\noindent
~\citeauthor{ref:zeiler:2012}~\cite{ref:zeiler:2012} noted that the units of the update step as presented in Equation~\ref{eq:heuristics:gd:adadelta_part_5} do not match. The weight update vector should have the same hypothetical units as the weight vector. To fix the aforemention problem, another exponentially decaying average is defined in terms of squared weight updates as is presented in Equation~\ref{eq:heuristics:gd:adadelta_part_6} below.

>>>>

      \item \textbf{Section~\ref{sec:heuristics:what_is_a_heuristic}} provides background information on the origins and definition of the term \index{heuristic}\textit{heuristic}. It is shown that \index{heuristic}heuristics are a class of algorithms that are used to solve optimisation problems.

@@

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''.~\citeauthor{ref:romanycia:1985}~\cite{ref:romanycia:1985} provides a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \acf{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that, on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \index{meta-heuristic}meta-heuristics make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. Each of these are presented in detail in the following sections.

@@

Although gradient-based \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \acp{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most widely used. Gradient-based heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics and other second-order derivative methods such as Newtons' method. There are many variants of gradient-based \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acs{GD} framework called \acf{BP}.

@@

\noindent
In the context of training \textit{shallow} \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer to the input layer, updating the weights at the output and hidden layers. The algorithm that propagates the error signal backwards is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\mathcal{E}$, with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} describes the \acs{BP} process in two steps:

      \item \textbf{Backward Propagation}: During this phase the error signal is propagated backwards from the output layer, through the hidden layer, to the input layer of the \acs{FFNN}. Weights at the output and hidden layers are then adjusted as functions of the backpropagated error signal.
\end{itemize}

\noindent
The update step for \acs{GD} by means of \acs{BP} can be formulated as is shown in Equations~\eqref{eq:heuristics:gd:update_step_part_1}-\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_1} above, $w_{i}$ is the weight value at index $i$, $t$ is the timestep, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and timestep $t$. $\Delta w_{i}(t)$ is specified by Equation~\eqref{eq:heuristics:gd:update_step_part_2} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_2} above, $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function, relative to the weight at index $i$. The learning rate is an important hyper-parameter in gradient-based \index{heuristic}heuristics as it controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given by Equation~\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_3} above, $t_{p}$ refers to the target value for pattern $p$, $o_{p}$ refers to the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

Finally, assuming the use of \index{sigmoid}sigmoid activation function, the partial derivative of the activation function, relative to the net input signal is given in Equation~\eqref{eq:heuristics:gd:update_step_part_4} below.

\begin{equation}
      \label{eq:heuristics:gd:update_step_part_4}
      \frac{\partial f}{\partial net_{p}} = o_p(1 - o_{p})
\end{equation}

\noindent
Again, assuming the use of \acs{SSE} as the loss function, the pseudo-code implementation for the generic \acs{GD} algorithm is taken from~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:gd} below.

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the generic \acf{GD} heuristic.}
      \label{algo:heuristics:gd:gd}
      \begin{algorithmic}
            \State Initialise the \acs{FFNN} weights and biases, the learning rate $\eta$ and the timestep/epoch $T=0$;
            \While{stopping conditions are not met}
            \State Let $\epsilon_{T} = 0$;
            \For{each training pattern p}
            \State Calculate the output of the hidden and output layers (feedforward);
            \State Compute the error signals of the hidden and output layers;
            \State Adjust the weights of the output and hidden layers (backpropagate);
            \State $\epsilon_{T}$ += $[\epsilon_{T_{p}} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
            \EndFor
            \State $T += 1$;
            \EndWhile
      \end{algorithmic}
\end{algorithm}

\noindent
A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $g$, resulting in a simplified weight update step as is shown in Equation~\eqref{eq:heuristics:gd:sgd} below.

\begin{equation}
      \label{eq:heuristics:gd:sgd}
      \begin{split}
            w = w - \eta g
      \end{split}
\end{equation}

\noindent
In Equation~\eqref{eq:heuristics:gd:sgd} above, $w$ refers to the weight vector, $\eta$ refers to the learning rate as before and $g$ refers to the gradient of the error function relative to weight vector. Note that, vector notation and some subscripts related to layers, indices, timesteps and training patterns are omitted for convenience.

@@

This section shines light on the algorithmic implementation of \acs{BP} with specific context to stochastic and batch training. The implementation of \acs{GD} using stochastic training is referred to as \acf{SGD}.

\acs{SGD} was one of the first widely used heuristics to train \acs{FFNN}, however, it is not without flaws. With stochastic training, only one training pattern is presented at each iteration/epoch. As such, weight updates are done with high variance and noise~\cite{ref:ruder:2016}. An illustration of the the fluctuations caused by \acs{SGD} during training is given in Figure~\ref{fig:heuristics:gd:sgd}.

@@

A compromise to overshooting minima that \acs{SGD} suffers from, is to make use of mini-batches of training patterns where a small number of training patterns are presented to the \acs{FFNN} at once. This is referred to as \textit{mini-batch} training. Input patterns from mini-batches are approximations of the total population of training patterns. According to~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} this has two main advantages:

@@

      \item The learning rate that has been introduced so far is applied to all elements of the weight update vector. If the training data is sparse and the features have varying frequencies, an equal update to all weight elements is inefficient. Larger weight updates are required for less frequently occurring features. Inefficient updates of all elements in the weight vectors contribute to the credit assignment problem~\cite{ref:rumelhart:1986}, common to \acs{GD} variants.

@@

Research shows that \acs{SGD} has difficulty navigating ravines~\cite{ref:sutton:1986}. Ravines are areas where the surface curves much more steeply in one dimension than in another. These ravines are common around local minima. As such, \acs{SGD} is shown to oscillate across the slopes of the ravine while only making minor progress towards the local minima.

\Acs{Momentum} ~\cite{ref:qian:1999} is a variant of the \acs{SGD} that helps accelerate \acs{SGD} in the relevant direction, dampening oscillations.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$, of the weight update vector of the previous timestep to the current weight update vector. An illustration of \acs{SGD} with and without momentum is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

@@

\noindent
The accumulation of momentum is presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1} below, while the update step is then amended in Equation~\eqref{eq:heuristics:gd:momentum_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_1}
      \begin{split}
            v = \alpha v - \eta g
      \end{split}
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_2}
      \begin{split}
            w = w + v
      \end{split}
\end{equation}

@@

\Acf{NAG} is a variant of the \acs{Momentum} heuristic developed by~\citeauthor{ref:sutskever:2013-2}~\cite{ref:sutskever:2013-2} and is inspired by~\citeauthor{ref:nesterov:1983}'s~\cite{ref:nesterov:1983} work on optimising convex functions. \acs{NAG} provides an improvement to the momentum accumulation term by providing a look-ahead term that better refines the weight update step.

@@

\noindent
The difference in the update step for \acs{Momentum} and the update step for \acs{NAG} as developed by~\citeauthor{ref:sutskever:2013-2}~\cite{ref:sutskever:2013-2} is described by~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} as follows. \acs{Momentum} first calculates the current gradient, represented by the small blue vector in Figure~\ref{fig:heuristics:gd:nag} and then takes a large step in the direction of the updated accumulated gradient, presented by the big blue vector. In contrast, \acs{NAG} first takes a big step in the direction of the previous accumulated gradient presented by the brown vector. At this point the gradient is measured and \acs{NAG} makes a correction represented by the red vector. The complete update step is represented by the green vector. By anticipating approximate future positions of the weights, the weight update step as defined by \acs{NAG} controls the optimisation process from going too fast and results in increased responsiveness~\cite{ref:bengio:2013}. Along with the same velocity update step for \acs{Momentum}, as presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1}, the \acs{NAG} weight update step is given in Equation~\eqref{eq:heuristics:gd:nag_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:nag_part_2}
      \begin{split}
            w = w + \alpha v - \eta g
      \end{split}
\end{equation}

@@

\Acf{Adagrad} is an adaptation of \acs{SGD} which implements a learning rate for every element in the weight vector and is developed by Duchi et al.~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \acs{Adagrad} adapts the learning rate to the elements of the weight vector, by  performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason, \acs{Adagrad} is well suited for dealing with situations where training data is sparse.

In the \acs{GD} variants presented thus far, the weight updates have been applied by making use of the same learning rate $\eta$, for all elements of the weight update vector. \Acs{Adagrad} uses a different learning rate for every weight element $w_{i}$, at every timestep $t$. The weight update step for \acs{Adagrad} is presented in Equation~\eqref{eq:heuristics:gd:adagrad_part_1} below.

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_1}
      \begin{split}
            w_{t+1,i} = w_{t,i} - \frac{\eta}{\sqrt{G_{t,ii} + \epsilon}}.g_{t,i}
      \end{split}
\end{equation}

\noindent
In Equation~\eqref{eq:heuristics:gd:adagrad_part_1} above, $w_{i}$ is the $i$-th element of the weight vector, $G_{t,ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$, is the sum of the squared gradients w.r.t. $w_{i}$, up to timestep $t$ and $\epsilon$ is a smoothing term that avoids division by zero and is set to an insignificant value such as $1 \times 10^{-8}$. Since $G_{t}$ contains the sum of the squares of the gradients along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product, represented by $\odot$, between $G_{t}$ and $g_{t,i}$. The simplified update step as implemented by \acs{Adagrad} is given in Equation~\eqref{eq:heuristics:gd:adagrad_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_2}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{\sqrt{G_{t} + \epsilon}} \odot g_{t,i}
      \end{split}
\end{equation}

\noindent
\Ac{Adagrad}'s main benefit is that it does not require manual tuning of the learning rate $\eta$. However, a problem with \acs{Adagrad} is in the accumulation of squared gradients in the denominator, represented by $G_{t}$. Since every term that is added is positive, $G_{t}$ keeps growing, leading to a situation where the learning rate shrinks to the point that the heuristic is no longer able to learn.

@@

Zeiler~\cite{ref:zeiler:2012} presents \acs{Adadelta}, an improvement of \acs{Adagrad} that eliminates the accumulation of squared gradients in the denominator.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \acs{Adagrad}, \acs{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average of squared gradients at timestep $t$, denoted by $E[g^{2}]_{t}$  depends on a fraction $\alpha$ of the previous average of squared gradients and the current gradient, similar to the update step for \acs{Momentum}. Also similar to \acs{Momentum}, $\alpha$ is set to 0.9. The update step for $E[g^{2}]_{t}$ is presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_1} below.

@@

\noindent
To illustrate the difference between \acs{Adagrad} and \acs{Adadelta}, the \acs{SGD} weight update step for \acs{Adagrad}, as presented in~\eqref{eq:heuristics:gd:adagrad_part_2} is rewritten such that the diagonal matrix $G_{t}$, is replaced with the moving average of squared gradients as presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_2}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{\sqrt{E[g^{2}]_{t} + \epsilon}} g_{t}
      \end{split}
\end{equation}

\noindent
Since the denominator is the \acf{RMS} error criterion of the gradient, the criteria short-hand notation is used as is depicted in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_3}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{RMS[g]_{t}} g_{t}
      \end{split}
\end{equation}

\noindent
~\citeauthor{ref:zeiler:2012}~\cite{ref:zeiler:2012} noted that the units of the update step as presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} do not match. The weight update vector should have the same hypothetical units as the weight vector. To fix the aforemention problem, another exponentially decaying average is defined in terms of squared weight updates as is presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_4} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_4}
      \begin{split}
            E[\Delta w^{2}]_{t} = \alpha E[\Delta w^{2}]_{t - 1} + (1 - \alpha)\Delta w_{t}^{2}
      \end{split}
\end{equation}

\noindent
The \acf{RMS} error of weight updates is given in Equation~\eqref{eq:heuristics:gd:adadelta_part_5} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_5}
      \begin{split}
            RMS[\Delta w]_{t} = \sqrt{E[\Delta w^{2}]_{t} + \epsilon}
      \end{split}
\end{equation}

\noindent
Note that $RMS[\Delta w]_{t}$ is unknown at timestep $t$. $RMS[\Delta w]_{t}$ is approximated with the \acf{RMS} of weight updates until the previous timestep. The learning rate in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} is then replaced. The weight update step for \acs{Adadelta} is then concluded in Equation~\eqref{eq:heuristics:gd:adadelta_part_6} below.
